---
title: "SPAROscore: Getting Started"
package: SPAROscore
author: Venkatesh Kamaraj
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPAROscore: Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
# load the necessary libraries
library(SPAROscore)
library(airway)
library(ggplot2)
```

# What is SPAROscore?
SPAROscore is a gene signature scoring method designed to be 
robust across diverse gene expression datasets.
SPAROscore adapts to varying levels of sparsity, allowing signature scores to be 
efficiently computed across bulk, single-cell, 
and spatial transcriptomic datasets.

The resulting scores quantify the relative expression of a gene signature 
compared with the background expression of the same sample/cell/domain, 
making the scores straightforward to interpret biologically.

Internally, SPAROscore ranks genes within each column and 
computes the Spearman footrule distance between the observed ranks of the 
signature genes and a background gene expression rank estimated from the 
geometric mean expression of the sample, cell, or spatial domain.


# How to use SPAROscore?
To use SPAROscore, at minimum, a gene expression dataset and
one or more gene signatures are required. 

To demonstrate this, we will use the gene expression matrix 
from the `airway` package, containing RNA-seq read counts from eight 
human airway smooth muscle (ASM) cell line samples: 
four treated with dexamethasone and four untreated controls.

As an example gene signature, we use the top differentially expressed genes 
reported in Himes et al.

## Load the dataset
```{r}
# load the airway dataset
data(airway)

# extract the gene expression matrix
counts_data <- assay(airway)

# extract the metadata
metadata <- as.data.frame(colData(airway))

# view sample level dexamethasone treatment status
metadata[, c("Run", "dex")]

# set the glucocorticoid response markers as the gene signature 
signature_genes <- c("ENSG00000112936", "ENSG00000170606", "ENSG00000120129", 
                     "ENSG00000152795", "ENSG00000211445", "ENSG00000163884",
                     "ENSG00000189221", "ENSG00000101347", "ENSG00000196136", 
                     "ENSG00000152583", "ENSG00000120658", "ENSG00000157514",
                     "ENSG00000103196", "ENSG00000179094", "ENSG00000152763",
                     "ENSG00000103310", "ENSG00000127954")
```

## Compute signature scores using SPAROscore
```{r, results = "markup"}
scores <- sparoscore(data = counts_data, signatures = signature_genes)

# view the scores
scores
```


## Visualise the results
Because the selected signature consists of glucocorticoid response markers, 
dexamethasone-treated ASM cell lines are expected to exhibit higher signature 
scores than untreated controls.
```{r}
# add the signature scores to the metadata
metadata$scores <- scores[, 1]

# plot results
ggplot(metadata, aes(x = dex, y = scores, fill = dex)) +
    geom_boxplot() +
    geom_jitter(width = 0.05) +
    xlab("Dexamethasone treatment status") +
    ylab("SPAROscore")+
    ggtitle("SPAROscore signature scores", 
            subtitle = "Stratified by treatment")+
    theme_classic()
```


# What transcriptomics technologies can SPAROscore handle?
SPAROscore can be applied to bulk RNA-seq, single-cell RNA-seq, and 
spatial transcriptomics data, 
adapting automatically to differences in data sparsity.

More information on transcriptomics technologies can be found here: 
[Transcriptomics technologies](sparoscore_transcriptomics_technologies.html)

# What data structures can SPAROscore handle?
SPAROscore accepts a wide variety of commonly used data structures, including
dense matrices, sparse matrices, DelayedArray, SummarizedExperiment, 
SingleCellExperiment, SpatialExperiment and Seurat objects.

More information on data structures can be found here:
[Data structures](sparoscore_data_structures.html)

# What are some advanced capabilities of SPAROscore?
To override the default approximation of background gene expression with 
the geometric averages, users can input custom values to be used as background
expression. Moreover, custom rank capping can also be performed by the user.

As gene rankings is a compute-intensive step, SPAROscore allows for storage of 
gene ranking for future scoring tasks.

For additional options and advanced usage, see:
[Advanced scoring options](sparoscore_advanced_options.html)


```{r}
sessionInfo()
```
