---
title: "SPAROscore: Getting Started"
author: "Venkatesh Kamaraj"
format: html
---

# What is SPAROscore?
SPAROscore is a gene signature scoring method designed to be 
robust across diverse gene expression datasets.
SPAROscore adapts to varying levels of sparsity, allowing signature scores to be 
efficiently computed across bulk, single-cell, 
and spatial transcriptomic datasets.

The resulting scores quantify the relative expression of a gene signature 
compared with the background expression of the same sample/cell/domain, 
making the scores straightforward to interpret biologically.

Internally, SPAROscore ranks genes within each column and 
computes the Spearman footrule distance between the observed ranks of the 
signature genes and a background gene expression rank estimated from the 
geometric mean expression of the sample, cell, or spatial domain.

# How to use SPAROscore?
To use SPAROscore, at minimum, a gene expression matrix and
one or more gene signatures are required. 

To demonstrate this, we will use the gene expression matrix 
from the `airway` package, containing RNA-seq read counts from eight 
human airway smooth muscle (ASM) cell line samples: 
four treated with dexamethasone and four untreated controls.

As an example gene signature, we use the top differentially expressed genes 
reported in Himes et al.

## Load the dataset
```{r, warning=FALSE, message=FALSE}
library(SPAROscore)
library(airway)
library(ggplot2)

# load the airway dataset
data(airway)

# extract the gene expression matrix
counts_data <- assay(airway)

# extract the metadata
metadata <- as.data.frame(colData(airway))

# set the glucocorticoid response markers as the gene signature 
signature_genes <- c("ENSG00000112936", "ENSG00000170606", "ENSG00000120129", 
                     "ENSG00000152795", "ENSG00000211445", "ENSG00000163884",
                     "ENSG00000189221", "ENSG00000101347", "ENSG00000196136", 
                     "ENSG00000152583", "ENSG00000120658", "ENSG00000157514",
                     "ENSG00000103196", "ENSG00000179094", "ENSG00000152763",
                     "ENSG00000103310", "ENSG00000127954")
```

## Compute signature scores using SPAROscore
```{r}
scores <- sparoscore(data = counts_data, signatures = signature_genes)

# view the scores
scores
```


## Visualise the results
Because the selected signature consists of glucocorticoid response markers, 
dexamethasone-treated ASM cell lines are expected to exhibit higher signature 
scores than untreated controls.
```{r}
# add the signature scores to the metadata
metadata$scores <- scores[, 1]

# plot results
ggplot(metadata, aes(x = dex, y = scores, fill = dex)) +
    geom_boxplot() +
    geom_jitter(width = 0.05) +
    xlab("Dexamethasone treatment status") +
    ylab("SPAROscore")+
    ggtitle("SPAROscore signature scores", 
            subtitle = "Stratified by treatment")+
    theme_classic()
```


# When to use SPAROscore?
SPAROscore can be applied to bulk RNA-seq, single-cell RNA-seq, and 
spatial transcriptomics data, 
adapting automatically to differences in data sparsity.

More information on transcriptomics technologies can be found here: 
[Transcriptomics technologies](sparoscore_transcriptomics_technologies.html)

# What data structures can SPAROscore handle?
SPAROscore accepts a wide variety of commonly used data structures, including
dense matrices, sparse matrices, DelayedArray, SummarizedExperiment, 
SingleCellExperiment, SpatialExperiment and Seurat objects

More information on data structures can be found here:
[Data structures supported by SPAROscore](sparoscore_data_structures.html)

For additional options and advanced usage, see:
[Advanced scoring options](sparoscore_advanced_options.html)


```{r}
sessionInfo()
```
